---
title: "An afrimapr intro to sf"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## An introduction to geographic data with package sf

Geographic data can be divided into :

* Vector
  * **points** 
  * **lines** 
  * **polygons**
  
* Raster
  * **gridded data** (e.g. from remote sensing)

The sf package deals with vector data (points, lines and polygons)


### Sources of geographic data in R

1. existing packages
1. local files (that we may have downloaded)
1. internet
1. created by hand in R


### Loading some polygon data

*This shows the code to load some polygon data from an existing package*

Now write a function that adds any two numbers and then call it:

```{r polygons-from-rnaturalearth, exercise=TRUE, exercise.lines = 5}

library(rnaturalearth)
library(sf)

africa_map <- ne_countries(continent = 'Africa', returnclass = 'sf')

plot(africa_map["pop_est"])
  

```




## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
