---
title: "afrimapr intro to spatial data in R"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  An entry level introduction to spatial data in R using examples from Africa.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

This is an entry level introduction to spatial data in R using examples from Africa.

It is aimed at those with a little knowledge of R.

## Spatial data

We will be looking at the following spatial data for Africa. 

1. Capital city locations (points)
1. A highway network (lines)
1. Country boundaries (polygons)
1. Population density (gridded or raster data)

The first 3 data types are termed **vector data** and the final one is **raster data**.

In R the `sf` package deals with vector data (points, lines and polygons), and the `raster` package deals with raster data.

There are other packages too but we don't need those for now.


## Loading packages and data

Packages in R contain extra methods and data not available in base R.

We will be loading one package (`afrilearndata`) containing example data for us to look at.

And the packages `sf` and `raster` allowing us to deal with vector and raster data.

To use an R package requires a 2 step process :
1. installation               needed infrequently from the internet
1. library([package_name])    needed each time you start a new R session

A package may have been installed already on your system because it is a 'dependency' needed for another package.

In this case the packages `afrilearndata`, `sf` and `raster` should already have been installed when you installed this package.

To check that the packages have been installed try running the `library` commands below :

```{r library_packages, exercise = TRUE, exercise.eval = FALSE, message = FALSE, warning = FALSE}

# for vector data handling
library(sf) 

# for raster data handling
library(raster) 

# example spatial data for Africa
library(afrilearndata)

```

If you happen to get messages indicating any of the packages are not installed you can use the code below to install them.

TODO do I want to put installation code in ? try to avoid danger that learners run it anyway when it is not needed ?


### 

An sf object is a dataframe with one row per feature and a column called geometry that contains the spatial information.

Use names() to see what columns there are. Then try changing the plot to show data from a different column.

```{r sf-object, exercise=TRUE, exercise.lines = 5}

library(rnaturalearth)
library(sf)

sf_africa <- ne_countries(continent = 'Africa', returnclass = 'sf')

names(sf_africa)

plot(sf_africa["pop_est"])

```






### FROM HERE OLD STUFF MAYBE TO RE_USE

### Other sources of geographic data in R

1. existing packages
1. local files (that we may have downloaded)
1. internet
1. created by hand in R

### polygon data from an existing package

*code to load some polygon data from an existing package*

```{r polygons-from-rnaturalearth, exercise=TRUE, exercise.lines = 5}

library(rnaturalearth)
library(sf)

sf_africa <- ne_countries(continent = 'Africa', returnclass = 'sf')

plot(sf_africa["pop_est"])
  

```


[NOTE can have setup and exercise chunks]
[https://github.com/rstudio/learnr/issues/72]
[NOTE can use exercise.setup='setupA' to make chunks dependent on previous setup or exercise chunks ]


## next Topic

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
