---
title: "afrimapr intro to spatial data in R"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  An entry level introduction to spatial data in R using examples from Africa.
---

```{r setup, include=FALSE}
library(learnr)

# default options for all chunks
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE )

# working with vector data
library(sf) 

# data wrangling
library(dplyr) 

# interactive mapping
library(mapview) 

# example data
library(afrilearndata)

# static and interactive mapping
library(tmap)

# for raster manipulation
library(raster)

```

This is an entry level introduction to spatial data in R using examples from Africa. It is aimed at those with a little knowledge of R.


## Spatial data outline

We will start by looking at these spatial data for Africa, shown in the map below. 

1. Capital city locations (points)
1. A highway network (lines)
1. Country boundaries (polygons)
1. Population density (gridded or raster data)

```{r tmap-first-view, echo=FALSE}
#echo=FALSE to not show code in first version 

# tmap_mode("view") to set to tmap interactive viewing mode

# getting error on deploying to shinyapps
#Error in value[3L] :OGRCreateCoordinateTransformation() returned NULL: PROJ available?
#potential solution
#https://stackoverflow.com/questions/61286108/error-in-cpl-transformx-crs-aoi-pipeline-reverse-ogrcreatecoordinatetrans
#If you save an sf-dataframe with a newer version of GDAL, and then try on a system with an older version of GDAL, the projection info cannot be read properly.
#The solution is to re-set the projection:
#st_crs(data) <- 4326

sf::st_crs(africapitals) <- 4326
sf::st_crs(afrihighway) <- 4326
sf::st_crs(africountries) <- 4326
sf::st_crs(africontinent) <- 4326 #not used yet but just in case

#not the raster layer
#st_crs(afripop2020) <- 4326

tmap::tm_shape(afripop2020) +
    tm_raster(palette = rev(viridisLite::magma(5)), breaks=c(0,2,20,200,2000,25000)) +
tm_shape(africountries) +
    tm_borders("white", lwd = .5) +
    #tm_text("iso_a3", size = "AREA") +
tm_shape(afrihighway) +
    tm_lines(col = "red") + 
tm_shape(africapitals) +
    tm_symbols(col = "blue", alpha=0.4, scale = .6 )+
tm_legend(show = FALSE)
```


Cities, highways and boundaries are examples of point, line and polygon data termed **vector data**. 

The gridded population density data are termed **raster data**.

In R there is often more than one package that does the same thing. Which one is 'best' for you may depend on preference and can change over time. This is true for R spatial operations.  

In R the `sf` package deals with vector data (points, lines and polygons), and the `raster` package deals with raster data.

There are other packages too but we don't need those for now.


## Loading packages and data

Packages in R contain extra methods and data to add to base R.

We will be loading one package (`afrilearndata`) containing example data for us to look at.

And the packages `sf` and `raster` allowing us to deal with vector and raster data.

To use an R package requires a 2 step process :
1. installation               needed infrequently from the internet
1. library([package_name])    needed each time you start a new R session

A package may have been installed already on your system because it is a 'dependency' needed for another package.

In this case the packages `afrilearndata`, `sf` and `raster` should already have been installed when you installed this package.

To check that the packages have been installed try running the `library` commands below. If they have been installed nothing should happen. Not very interesting but a good check and a good reminder that this is what you will need to do each time you start a new R session.

```{r library_packages, exercise = TRUE, exercise.eval = FALSE, message = FALSE, warning = FALSE}

# for vector data handling
library(sf) 

# for raster data handling
library(raster) 

# example spatial data for Africa
library(afrilearndata)

```

If you happen to get messages indicating any of the packages are not installed you can use `install.packages` to install them.

[TODO do I want to put installation code in ? try to avoid danger that learners run it anyway when it is not needed ?]


## First look at spatial data objects 

We are going to take a look at the spatial data objects used to create the map shown at the start of the tutorial.

We call them 'objects' because the data are already stored in R. This is also to make clear the difference from a 'file' that is stored elsewhere on your computer. A 'file' can be read into an R 'object' and we will come to that later.  

First we will look at capital cities which are stored in an object called `africapitals`.

Using the plot method should display a number of maps of the point locations of African capitals. 

```{r sf-points-plot, exercise=TRUE, exercise.lines = 5}

plot(africapitals)
#plot(sf::st_geometry(africapitals))

```

This uses the plot function defined in the `sf` package. It creates a series of maps in each one the points are coloured according to the values stored in one column. There is a function in sf called `st_geometry()` that allows you to get just the spatial parts without the attributes and you can see the result of that by removing the # in front of the 2nd line of code in the window above and pressing 'Run code'.  But we are getting ahead of ourselves, let us have a look at the structure of the object itself.

In R there are various functions that can help us explore what an object contains. We find these particularly useful, there is some overlap between them.

1. str()        structure of the object, displays both names and values
1. head()       displays the first values and their names
1. names()      gives just column names
1. class()      gives the class of the object, broadly what sort of object it is  

Have a look at the outputs for `africapitals` :

```{r sf-points-str, exercise=TRUE, exercise.lines = 5}

str(africapitals)
head(africapitals)
names(africapitals)
class(africapitals)

```

These show us that `africapitals` is of class `sf` and `data.frame` and contains a series of columns including ones named : 'capitalname', 'countryname' and 'geometry'.

`data.frame`, often referred to as just dataframe, is the most common object type in R certainly for new users. Dataframes store data in rows and named columns like a spreadsheet. 

`sf` objects are a special type dataframe with a column called 'geometry' that contains the spatial information, and one row per feature. In this case the features are points.

If you look at the output from the `str()` command above you should see that the first value in the geometry column has the coordinates 7.17 9.18. Because the capitals data are points they just have a single coordinate pair representing the longitude and latitude of each capital.







## ...


### FROM HERE OLD STUFF TO RE_USE

Use names() to see what columns there are. Then try changing the plot to show data from a different column.

### Other sources of geographic data in R

1. existing packages
1. local files (that we may have downloaded)
1. internet
1. created by hand in R

### polygon data from an existing package

*code to load some polygon data from an existing package*

```{r polygons-from-rnaturalearth, exercise=TRUE, exercise.lines = 5}

  

```


[NOTE can have setup and exercise chunks]
[https://github.com/rstudio/learnr/issues/72]
[NOTE can use exercise.setup='setupA' to make chunks dependent on previous setup or exercise chunks ]


### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
