---
title: "How to get your own spatial data into R"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  An entry level outline to help you get your own data, of different types, into R as a first step to mapping it.
---

```{r setup, include=FALSE}
library(learnr)

# default options for all chunks
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE )

library(sf)               # working with vector data
library(afrilearndata)    # example data
library(tmap)             # static and interactive mapping
library(raster)           # raster manipulation
#library(dplyr)           # data wrangling
#library(mapview)         # interactive mapping

```

## Intro to this tutorial & afrilearnr

This entry level tutorial will demonstrate getting spatial data of different types into R. The aim is to support you getting your own data into R as a first step to making maps or other plots.

The [afrilearnr](https://github.com/afrimapr/afrilearnr) package contains tutorials to teach spatial in R with African data. It is part of the [afrimapr](https://afrimapr.github.io/afrimapr.website/) project.

Through the magic of [learnr](https://rstudio.github.io/learnr/) you can modify the R code in the boxes below and press run to see results.

If you are accessing this from shinyapps you can also install [afrilearnr from github](https://github.com/afrimapr/afrilearnr) and run these tutorials locally.


## csv or xls file with coordinates

Show two stage process of reading in a csv and then converting to sf.
Demo what happens if don't set the crs (i.e. mapview can't put it on world map)
Talk about coords
Talk about 4326

[https://ourairports.com/continents/AF/airports.csv]

```{r csv-airports, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

#read from a csv file stored in the afrilearndata package as example to work with
dfairports <- readr::read_csv(system.file("extdata/afriairports.csv", package="afrilearndata"))

# to convert to sf object
afriairports <- sf::st_as_sf(dfairports, coords=c("longitude_deg", "latitude_deg"), crs=4326)

# mapview for a quick interactive plot
mapview(afriairports)

# making mapview plot more informative
mapview(afriairports, zcol='type', label='name', cex=2)

```


## shapefile (*.shp)


## raster tiff 
