---
title: "How to get your own spatial data into R"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  An entry level outline to help you get your own data, of different types, into R as a first step to mapping it.
---

```{r setup, include=FALSE}
library(learnr)

# default options for all chunks
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE )

library(sf)               # working with vector data
library(afrilearndata)    # example data
#library(tmap)             # static and interactive mapping
library(raster)           # raster manipulation
#library(dplyr)           # data wrangling
library(mapview)          # interactive mapping
library(readr)            # reading text files

library(rgdal)            # to try to correct raster error on shinyapps

```

## Intro to this tutorial & afrilearnr

This entry level tutorial will demonstrate getting spatial data of different types into R. The aim is to support you getting your own data into R as a first step to making maps or other plots.

The [afrilearnr](https://github.com/afrimapr/afrilearnr) package contains tutorials to teach spatial in R with African data. It is part of the [afrimapr](https://afrimapr.github.io/afrimapr.website/) project.

Through the magic of [learnr](https://rstudio.github.io/learnr/) you can modify the R code in the boxes below and press run to see results.

If you are accessing this from shinyapps you can also install [afrilearnr from github](https://github.com/afrimapr/afrilearnr) and run these tutorials locally.

See the sections for different file types following. In each we will demonstrate getting the data into R and then plotting with the `mapview` package. `mapview` is a great option for plotting data quickly to check that it is how (and where!) you expect.

To repeat these steps locally you would need the following packages.

```{r packages, echo=TRUE}

library(readr)            # reading text files
library(sf)               # working with vector data
library(mapview)          # interactive mapping
library(raster)           # raster manipulation
library(afrilearndata)    # example data, not required if using only your own data

```

## .csv, .txt or .xls file with coordinates

Text files containing point data are one of the commonest file types that we see in small-scale operational mapping of data. Usually these consists of one row per record (e.g. the location of a health facility or disease case or dwelling) with two columns containing the coordinates of the location (e.g. longitude & latitude or x & y), and other columns containing attributes of that location (e.g. facility or disease type). 

These files can be `.csv` comma delimited, or `.txt` space delimited or various spreadsheet formats including `.xls`.

To map these data in R usually requires a 3 step process.

1. read the data file into an R dataframe
1. convert the dataframe into an R spatial (package `sf`) object
1. plot the `sf` object

Here we will demonstrate the 3 steps using some airport data from the excellent [ourairports](https://ourairports.com/continents/AF/airports.csv) that we have extracted and saved in the [afrilearndata](https://github.com/afrimapr/afrilearndata) package. 

```{r csv-airports, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

# 1. read into dataframe
filename <- system.file("extdata/afriairports.csv", package="afrilearndata")
mydf <- readr::read_csv(filename)

# 2. convert to sf object
mysf <- sf::st_as_sf(mydf, 
                     coords=c("longitude_deg", "latitude_deg"),
                     crs=4326)

# 3. quick interactive plot
mapview(mysf)    


```


To apply the code chunk above to your own data :

* replace `system.file("extdata/afriairports.csv", package="afrilearndata")` with the path to your file (this might just be something like `"myfile.csv"`) 
* replace `"longitude_deg", "latitude_deg"` with the names of the columns containing the coordinates in your data
* you may need to change `crs=4326` as explained below

### CRS
`crs` stands for Coordinate Reference System. It determines how coordinates are converted to a location on the Earth. 
In this case it tells `sf` what system to expect. In the majority of cases coordinates (e.g. collected from a GPS) are stored in a system represented by the code `4326`. 4326 is the EPSG code for longitude, latitude using the WGS84 datum, but you don't really need to know that. 4326 is a good number to remember !

Question : What happens when the `crs=4326` argument is not included in the code below ? Try adding it back in and see what the difference is.


```{r csv-crs-missing, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

# 1. read into dataframe
filename <- system.file("extdata/afriairports.csv", package="afrilearndata")
mydf <- readr::read_csv(filename)

# 2. convert to sf object - note crs missing
mysf <- sf::st_as_sf(mydf, 
                     coords=c("longitude_deg", "latitude_deg"))

# 3. quick interactive plot
mapview(mysf)    

```

Answer : You should see that when the `crs` argument is missed out the sf object is still created but mapview is unable to position it in the world so although the points appear there is no map background.

### .xls files

For Microsoft Excel files you just need to change step 1 of the three step approach. You can read an excel file into a dataframe using the package [readxl](https://readxl.tidyverse.org/) with something like `readxl::read_excel(filename)`. Another option is to save the sheet that you want as a .csv file from MS Excel itself.

### improving `mapview` maps

In these examples we use `mapview` to give us a quick view by passing it only the spatial object. `mapview` is very flexible and by passing just a few arguments the map can be made much more informative. Try pasting this line into the code window above and running it. It uses the columns named `type` and `name` that are in the datafile to colour and label the points.

`mapview(afriairports, zcol='type', label='name', cex=2)`


### R dataframe example to play with

You can also replace the first file reading step by directly creating a dataframe within R.

In the example below try changing the coordinates within the dataframe at step 1, and run to see the points change.

```{r dataframe-sf, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

# 1. create dataframe
mydf <- data.frame(x=c(-10,10,30),
                   y=c(20,0,-20),
                   attribute=c("a","b","c"))

# 2. convert to sf object
mysf <- sf::st_as_sf(mydf, 
                     coords=c("x", "y"),
                     crs=4326)

# 3. quick interactive plot
mapview(mysf)    

```


## shapefile (.shp)

Shapefiles continue to be a common format for spatial data despite the fact that they are rather old now and some things about them are not ideal. One thing that can confuse users is that a shapefile consists of a collection of files with the same name and different suffixes. If some of the files are not present then it may no longer be possible to get at the data.

e.g. myfile.shp, myfile.shx, myfile.dbf, myfile.prj

If a colleague emails use just a single file named `*.shp` then you will not be able to map it in R. You would need to ask them to email you all of the files.

Shapefiles can store points, lines or polygons. The example below uses a shapefile containing polygons.


```{r shp-countries, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

# read file into a spatial object
filename <- system.file("extdata/africountries.shp", package="afrilearndata")
africountries <- sf::read_sf(filename)

# quick interactive plot
mapview(africountries)

```

Because shapefiles are spatial files they can be read directly into a spatial (`sf`) object in R with `sf::read_sf(filename)`. This combines steps 1 & 2 from the csv example. In addition you don't need to specify in R which columns contain the coordinates or what the Coordinate Reference System (crs) is. This is effectively because these two steps will have been done when the file was created.


## .kml, .gpkg & .json

For other spatial vector formats (e.g. kml, geopackage & geojson) the same approach as for a shapefile usually works i.e. `sf::read_sf(filename)`.

Here we show an example with a .kml file of the simplified African highway network.

```{r kml-highways, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

filename <- system.file("extdata/trans-african-highway.kml", package="afrilearndata")

afrihighway <- sf::read_sf(filename)

# quick interactive plot
mapview(afrihighway)

```


## raster tiff 

TODO
* say similar to vector but uses raster package
* can still use mapview to view
* other raster formats

```{r rast1, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

filename <- system.file("extdata/afripop2020.tif", package="afrilearndata")

myrast <- raster::raster(filename)

# quick interactive plot
mapview(myrast)

```
