---
title: "A crash course in R mapping with African data"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  Crash course in making maps in R with minimal explanation.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Intro to afrilearnr & this crash course

The [afrilearnr](https://github.com/afrimapr/afrilearnr) package contains tutorials to teach spatial in R with African data. It is part of the [afrimapr](https://afrimapr.github.io/afrimapr.website/) project.

This is a crash course going straight to the endpoints of other tutorials. It provides the code to create maps that you can experiment with. It has minimal explanation. If you want more explanation consult the other tutorials.

Through the magic of [learnr](https://rstudio.github.io/learnr/) you can modify the R code in the boxes below and press run to see results.

Datasets are obtained from the [afrilearndata](https://github.com/afrimapr/afrilearndata) package that provides example point, line, polygon and raster data for Africa. 

If you are accessing this from shinyapps you can also install [afrilearnr from github](https://github.com/afrimapr/afrilearnr) and run these tutorials locally.


First loading required packages, if you see any warnings below you may need to install those you don't have.
```{r libraries, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

library(sf)     # for vector data handling
library(raster) # for raster data handling
library(tmap) # static & interactive mapping
library(mapview) #interactive mapping
library(afrilearndata) # example spatial data for Africa
library(rgdal) #seems to be needed for mapview raster

#you may need :
#remotes::install_github("afrimapr/afrilearnr")

#temporary fix for shinyapps & may help for users with an older version of rgdal
sf::st_crs(africapitals) <- 4326
sf::st_crs(afrihighway) <- 4326
sf::st_crs(africountries) <- 4326
sf::st_crs(africontinent) <- 4326 

```

## tmap : map vector and raster data

`tmap` uses static 'plot' mode by default. To make these plots interactive, copy & paste `tmap_mode('view')` above the code below & press `Run Code`.

```{r tmap-vector-raster, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE, exercise.setup='libraries'}

tmap::tm_shape(afripop2020) +
    tm_raster(palette = rev(viridisLite::magma(5)), breaks=c(0,2,20,200,2000,25000)) +
tm_shape(africountries) +
    tm_borders("white", lwd = .5) +
    #tm_text("iso_a3", size = "AREA") +
tm_shape(afrihighway) +
    tm_lines(col = "red") + 
tm_shape(africapitals) +
    tm_symbols(col = "blue", alpha=0.4, scale = .6 )+
tm_legend(show = FALSE)

```

## tmap : colour polygons by data

`tmap` uses static 'plot' mode by default. To make these plots interactive, copy & paste `tmap_mode('view')` above the code below & press `Run Code`.

```{r tmap-colour-polygons, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE, exercise.setup='libraries'}


tm_shape(africountries) +
    tm_polygons("pop_est")
    #try commenting out above line and uncommenting that below
    #tm_polygons("income_grp")

```

## ggplot : colour polygons by data

`ggrepel` helps avoid label overlap if plot is big enough, reduce label `size` to fit in smaller plots.
Comment out `geom_text_repel` lines to remove labels.
Comment out `theme_void()` to add axes back in.

```{r ggplot-colour-polygons, exercise = TRUE, exercise.eval = TRUE, warning = FALSE, exercise.setup='libraries'}

library(ggplot2)
library(ggrepel)

ggplot(africountries) +
    geom_sf(aes(fill = pop_est)) +
    scale_fill_viridis_c() +
    theme_void() +
    geom_text_repel(aes(label=name_long, geometry=geometry),
                    stat="sf_coordinates",
                    point.padding = NA, #allows points to overlap centroid
                    colour='darkgrey', size=3
                   ) +
    labs(title = "Population by country 2000", fill = "Population Estimate")

```


## mapview : interactive point data

```{r mapview-points, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE, exercise.setup='libraries'}

mapview(africapitals,
        zcol = 'pop',
        #cex = 'pop', #to size points by attribute
        label = paste(africapitals$countryname,':',africapitals$capitalname))

```

## mapview : interactive points and lines

Using `list(1,2)` to pass two layers and respective arguments.

```{r mapview-points-lines, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE, exercise.setup='libraries'}

mapview(list(africapitals, afrihighway),
        zcol = list('pop','Name'),
        legend = list(TRUE, FALSE),  
        color = list(NULL, 'red'),
        label = list(paste(africapitals$countryname,':',africapitals$capitalname),
                     afrihighway$Name) )

```

## mapview : interactive raster

Rasters are displayed semi-transparent by default so base layers including place names are visible. Population density data are often highly skewed with rare high values meaning that most variation gets hidden when using default data breaks between colours. Specifying near logarithmic breaks using `at` below makes patterns clearer.

```{r mapview-raster, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE, exercise.setup='libraries'}

mapview(afripop2020,
        at = c(0,2,20,200,2000,25000))

```

Note that previously this failed on shinyapps : package 'rgdal' is not available. Why when the previous exercise worked is beyond me. Adding library(rgdal) at the start of this script fixed it. Strange because I would think that rgdal would be imported by mapview if needed.

## display one country

We can select one row from `africountries` in `afrilearndata` to allow us to display a country, using the `bbox` argument in tmap, the `ext` argument allows us to extend the bbox a little.


```{r mapview-one-country, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE, exercise.setup='libraries'}

tmap_mode('view') #tmap uses static 'plot' mode by default, uncomment this to change to interactive 'view' mode
#tmap_mode('plot') #to change back to static 'plot' mode

country_name <- "Mozambique" #"Botswana" # names need to match exactly, can also use name_fr etc.

sfcountry <- dplyr::filter(africountries, name==country_name)

tmap::tm_shape(afripop2020, bbox=sf::st_bbox(sfcountry), ext=1.1) +
    tm_raster(palette = rev(viridisLite::viridis(5)), breaks=c(0,2,20,200,2000,25000)) +
    tm_layout(main.title=country_name, legend.outside=TRUE) +
tm_shape(africountries) +
    tm_borders("grey", lwd = 2) 

```



[TODO add osmdata]


